\chapter{Editor's Preface}
\label{ch:editor}

\begin{fullwidth}
This book is Raymond Cottrell's paper \textit{The ``Sanctuary Doctrine''---Asset 
or Liability}, laid out to the design of Edward Tufte's books.  

The paper was first delivered to the second JIF symposium, November 2 through 
November 4, 2001, and again publicly on February 9, 2002 at the Association of 
Adventist Forums meeting in San Diego, California.

I do not have access to the original paper, so this text is based on digital copies
located at
\begin{itemize}
    \item \url{http://www.rethinkingadventism.com/support-files/cottrell_1844.pdf}, and
    \item \url{http://www.truthorfables.com/Sanctuary_Cottrell.htm}.
\end{itemize}

This layout incorporates the following design decisions in the conversion of
the original paper's footnotes to sidenotes:
\begin{enumerate}
    \item Extra page breaks prevent sidenotes from extending past the 
    length of the page. As a result, some pages contain a small amount of body
    text, but a large amount of sidenote text.
    \item Appendix chapter~\nameref{ch:shp} replaces footnote 131.
    \item For each footnote that refers to a passages of scripture, I include
    the text of the passage within the corresponding sidenote.  Exceptions
    to this are footnotes that refer to Leviticus 16, to Matthew 24 and 25,
    Daniel 2 and Daniel 7--12. I include the complete chapters for each
    of these in the Appendix.
    \item For those footnotes that are used in multiple locations, directly or 
    indirectly, I duplicate the contents of those footnotes in 
    multiple sidenotes, rather than having subsequent sidenotes refer to
    previous sidenotes. For this reason, the sidenote numbers in this document
    do not always match the footnote numbers from the source material.
    An exception to this rule is when the original
    footnote and its ``duplicate'' appear consecutively in the text.
\end{enumerate}

In the source material for this layout, footnote 37 reads ``see note 26''.
However, footnote 26 has nothing to do with the text for which footnote 37
appears.  I am unable to reconcile this
mismatch, so this document does not contain a sidenote for the original
footnote 37.



The majority of Cottrell's other papers referenced in this work are not
published.
\end{fullwidth}
